Charged pion emission from neutron-rich heavy ion collisions for studies on the symmetry energy

### 2020-08-14 MSU-thesis_estee_MaxFormat.tex
Submit to University

\sisetup{input-uncertainty-signs}

@@

Besides measuring neutrons stars directly, the laboratory provides the only way to study dense nuclear matter at different densities and proton-neutron asymmetries. It has been shown that charged pions are produced from nucleon-nucleon collisions through the short-lived intermediate $\Delta$ resonances, in the early, dense parts of Heavy Ion Collisions (HICs). Yet, existing pion experimental data tailored to answering questions about the symmetry energy was lacking and suffers from unknown systematic errors. This prompted the construction of the Samurai Pion-Reconstruction and Ion Tracker Time Projection Chamber (\spirit TPC) which was constructed to measure charged pions produced in neutron-rich HICs with a high efficiency. This Dissertation encompasses the first pion spectra from the most neutron-rich ($\tin{132}{124}$), and neutron-poor ($\tin{108}{112}$) systems measured at \SI{270}{\MeVA} taken at the RIBF facility at RIKEN. The total pion ratio was measured to within 4\% and the pion spectra are measured with a high degree of accuracy, marking the first time pion spectra  -- at energies relevant to the symmetry energy -- have been published.  Also, we will compare the results with the most recent transport models which can simulate pion production. We will highlight some of the successes and hope to motivate a discussion in the theoretical community on how to better reproduce the pion phenomena in neutron-rich HICs, which is important for not only constraining the density dependence of the symmetry energy, but also for understanding the $\Delta$ and pion roles in neutron stars. 
\end{abstract}

@@

\begin{dedication}
\centering
{\it
To my wife for her unchanging support and patience\\ through this journey,
and to my flowering bud, Hanna. 
}
\end{dedication}

@@

I am forever grateful to my mother and father for their continued support through my education. I am very blessed to have them as my parents. I wanted to thank my Aunts, Uncles, and Grandparents who have helped me financially and supported me throughout my schooling. To my Uncle Brian York, for his advice through the years and stimulating discussions.

>>>>

\hypersetup{draft}
\sisetup{input-uncertainty-signs}

@@

Besides observing neutrons stars directly, the laboratory provides the only way to study dense nuclear matter at different densities and proton-neutron asymmetries. It has been shown that charged pions are produced from nucleon-nucleon collisions through the short-lived intermediate $\Delta$ resonances, in the early, dense parts of Heavy Ion Collisions (HICs). Yet, existing pion experimental data tailored to answering questions about the symmetry energy was lacking and suffers from unknown systematic errors. This prompted the construction of the Samurai Pion-Reconstruction and Ion Tracker Time Projection Chamber (\spirit TPC) which was constructed to measure charged pions produced in neutron-rich HICs with a high efficiency. This Dissertation encompasses the first pion spectra from the most neutron-rich ($\tin{132}{124}$), and neutron-poor ($\tin{108}{112}$) systems measured at \SI{270}{\MeVA} taken at the RIBF facility at RIKEN. The total pion ratio was measured to within 4\% and the pion spectra are measured with a high degree of accuracy, marking the first time pion spectra  -- at energies relevant to the symmetry energy -- have been published.  Also, we will compare the results with the most recent transport models which can simulate pion production. We will highlight some of the successes and hope to motivate a discussion in the theoretical community on how to better reproduce the pion phenomena in neutron-rich HICs, which is important for not only constraining the density dependence of the symmetry energy, but also for understanding the $\Delta$ and pion roles in neutron stars. 
\end{abstract}

@@

\begin{dedication}
\centering
{\it
To my wife for her unwavering support and patience,\\
and to my flowering bud, Hanna. 
}
\end{dedication}

@@

I am forever grateful to my mother and father for their continued support through my education. I am very blessed to have them as my parents. I wanted to thank my Aunts, Uncles, and Grandparents who have helped me financially and supported me throughout my schooling. To my Uncle Brian York, for his advice through the years and his support throughout this journey.

### 2020-08-14 chapters_altFormat/ch_intro.tex
Submit to University

Guided by Eq.~\ref{eq:semiEmp}, we can separate the energy density $E$ of a system into two components,

@@

The pressure arising from the symmetry energy depends on the curvature of the symmetry energy at a given density as described in Eq.~\ref{eq:densEos}. If the density dependence of the symmetry energy grows with a positive derivative at high densities, the symmetry energy would in general force neutrons out of the high density system. If the symmetry energy decreases at high densities, making the derivative  negative, then the symmetry energy term would tend to attract neutrons. The symmetry energy is referred to as \emph{soft} if the second order term in the Taylor expansion -- at saturation density-- is negative, and \emph{stiff} in the case of a curvature that is close to zero or positive. It is the pressure coming from the symmetry energy that is driving the relative dynamics of neutrons and protons, where a stiff symmetry energy expels neutrons from dense matter and vice versa for the soft symmetry energy.

>>>>

Guided by Eq.~\ref{eq:semiEmp}, we can separate the energy per nucleon, $E$, of a system into two components,

@@

The pressure arising from the symmetry energy depends on the curvature of the symmetry energy at a given density as described in Eq.~\ref{eq:densEos}. If the density dependence of the symmetry energy grows with a positive derivative at high densities, the symmetry energy would in general force neutrons out of the high density system. If the symmetry energy decreases at high densities, making the derivative  negative, then the symmetry energy term would tend to attract neutrons. The symmetry energy is referred to as \emph{soft} if the second order term in the Taylor expansion -- at saturation density-- is negative, and \emph{stiff} in the case of a curvature that is close to zero or positive. It is the pressure coming from the symmetry energy that is driving the relative dynamics of neutrons and protons, where a stiff symmetry energy expels neutrons from dense matter and expels them less for a softer symmetry energy.

### 2020-08-14 chapters_altFormat/ch_results.tex
Submit to University

\begin{figure}[!htb]
\centering
\includegraphics[width=.8\textwidth]{singleRatio108_log.png}
\caption{Pion spectral ratio $\tin{108}{112}$ system and comparisons to 7 theoretical models.}
\label{fig:SRsn108}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=.8\textwidth]{singleRatio132_sum.png}
\caption{Pion spectral ratio $\tin{132}{124}$ system and comparisons to 7 theoretical models.}
\label{fig:SRsn132}
\end{figure}

>>>>

\begin{figure}[!htb]
\centering
\includegraphics[width=.8\textwidth]{singleRatio108_log.png}
\caption{Pion spectral ratio $\tin{108}{112}$ system and comparisons to 6 theoretical models.}
\label{fig:SRsn108}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=.8\textwidth]{singleRatio132_sum.png}
\caption{Pion spectral ratio $\tin{132}{124}$ system and comparisons to 6 theoretical models.}
\label{fig:SRsn132}
\end{figure}
